Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-03-31 sources/bmc_template_submission/bmc_article.tex
some edits

These sequences do appear in the Physcraper results in the list of matches from GenBank which did not fit the sequence length cutoffs set in the configuration. 
This ``seqlen\_mismatch.txt'' file includes the accession number, taxon, and sequence length of all sequences culled based on sequence length.

@@

Organellar genome sequences, such as chloroplasts and mitochondria will also generally be excluded from automatic addition based on length cutoffs.
Multiple sequence alignment of loci of drastically different lengths is unfeasible, and we have found in testing often returns incorrect results, splitting shorter sequences with many long gaps to align with exact matches across the entire longer locus. 
While it would be possible to directly extract the blast match from genomes, this match would exclude potentially homologous flanking regions which are not matched by blasts local search algorithm, but may be important for phylogenetic inference.
Instead we list the accession numbers for these matches in the seqlen\_mismatch file, for users to assess appropriate homologous regions to their alignment of interest.

Single gene phylogenies with very high numbers of taxa may lack sufficient signal for accurate resolution \cite{morel_phylogenetic_2020}.
The Physcraper workflow avoids some of these challenges, by focussing on ingroup taxa of an exisiting phylogeny, using markers that have been assessed and proven apporpraite for that phylogenetic scope in past publictions.
In addition, we thin alignments by removing identical sequences, and by setting a maximum number of sequences per taxon.
Nonetheless, it is encumbent on users to assess their final inference with respect to statistical support and biological plausibility.

Most of the default thresholds parameter values and software choices in the Physcraper pipeline are arbitrary. 
We hope that by making the process of locating homologous sequences online reproducible, 
and tracking what cutoffs are used, we make it easier for researchers to delve into the effect that different choices have on their inferences.
In addition, by providing the output files at each step of the analysis, 
it is straightforward to assess how changing parameter and software choices do or do not drive differences in phylogenetic inference.
By gathering the sequences, and making the unaligned files easily available to users, researchers can compare if applying any alternate alignment tool of their choice affects inferences. 
Once sequences are aligned, they can apply and compare inferences from any phylogenetics software.

@@

Physcraper generates gene trees, which individually do not capture the full complexity
of species' evolutionary history \cite{song2012resolving}.
However, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent
processes with ASTRAL \cite{mirarab2014astral} or SVD Quartets \cite{chifman2014quartet}).
Rigorous analyses of multiple loci allows for more complex evolutionary models
than analyses of large genomic data sets, and can provide better evolutionary estimates.
For example \cite{zhu_complexity_2021} show that when applying coalescent models, there is more information in two genes of 300 bp each than in 600 independent sites 
We have included a script (``multi_locus.py'') and examples of how to automatically merge the outputs of Physcraper runs at different loci into input files for ASTRAL, SVD Quartets, or as concatenated alignments for concatenated analyses.

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
These taxonomic links, and comparisons to existing published phylogenies in the OpenTree data store can also help flag paralogous sequences.
Accidentally including paralogs as homologs is a risk in phylogenetic analyses, and can be more prevalent in automated analyses than in manually curated analyses.
We provide users with several tools to try to assess homology of their aligned sequences. 
The estimated gene tree itself is an evolutionarily explicit way to visualize gene evolution, which in concert with taxonomic labelling can reveal paralogy.
Our conflict analysis tool informs the users of whether their tree contains major conflicts with established taxonomy. 
This conflict tool also returns information on if these taxonomic conflicts existed in the original input tree. 
These conflicts may be a sign that taxonomy needs to be updated, or may be a sign that non-homologous sequences have been included in the analysis. 
These taxonomic and phylogenetic conflicts flag regions for the tree for the researcher to more closely examine and assess homology.

@@

\section*{Conclusions}
Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible phylogenetics that has the
potential to consistently provide context for these ideas, highlighting the
importance of data sharing and open science in the field, biology and science.

>>>>

These sequences do appear in the Physcraper results in the list of matches from GenBank which did not fit the sequence length cutoffs set in the configuration.
This "seqlen\_mismatch.txt" file includes the accession number, taxon, and sequence length of all sequences filtered based on sequence length.

@@

Organellar genome sequences, such as chloroplasts and mitochondria will also generally be excluded from automatic addition based on length cutoffs.
Multiple sequence alignment of loci of drastically different lengths is unfeasible, and we have found in testing that it often returns incorrect results, splitting shorter sequences with many long gaps to align with exact matches across the entire longer locus.
While it would be possible to directly extract the blast match from genomes, this match would exclude potentially homologous flanking regions which are not matched by blasts local search algorithm, but may be important for phylogenetic inference.
Instead we list the accession numbers for these matches in the "seqlen\_mismatch.txt" file, for users to assess appropriate homologous regions for their alignment of interest.

Single gene phylogenies with very high numbers of taxa may lack sufficient signal for accurate resolution \cite{morel_phylogenetic_2020}.
The Physcraper workflow avoids some of these challenges, by focussing on ingroup taxa of an existing phylogeny, using markers that have been assessed and proven apporpraite for that phylogenetic scope in past publictions.
In addition, Physcraper thins alignments by removing identical sequences, and by setting a maximum number of sequences per taxon.
Nonetheless, it is incumbent on users to assess their final inference with respect to statistical support and biological plausibility.

Most of the default thresholds parameter values and software choices in the Physcraper pipeline are arbitrary.
We hope that by making the process of locating homologous sequences online reproducible,
and tracking what cutoffs are used, we make it easier for researchers to delve into the effect that different choices have on their inferences.
In addition, by providing the output files at each step of the analysis,
it is straightforward to assess how changing parameter and software choices do or do not drive differences in phylogenetic inference.
By gathering the sequences, and making the unaligned files easily available to users, researchers can compare if applying any alternate alignment tool of their choice affects inferences.
Once sequences are aligned, they can apply and compare inferences from any Phylogenetic software.

@@

Physcraper generates gene trees, which individually do not capture the full complexity
of species' evolutionary history \cite{song2012resolving}.
However, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent
processes with ASTRAL \cite{mirarab2014astral} or SVD Quartets \cite{chifman2014quartet}).
Rigorous analyses of multiple loci allows for more complex evolutionary models
than analyses of large genomic data sets, and can provide better evolutionary estimates.
For example \cite{zhu_complexity_2021} show that when applying coalescent models, there is more information in two genes of 300 bp each than in 600 independent sites
We have included a script ("multi_locus.py") and examples of how to automatically merge the outputs of Physcraper runs at different loci into input files for ASTRAL, SVD Quartets, or as concatenated alignments for concatenated analyses.

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
Taxonomic links, and comparisons to existing published phylogenies in the OpenTree data store can also help flag paralogous sequences.
Accidentally including paralogs as homologs is a risk in phylogenetic analyses, and can be more prevalent in automated analyses than in manually curated analyses.
We provide users with several tools to try to assess homology of their aligned sequences.
The estimated gene tree itself is an evolutionarily explicit way to visualize gene evolution, which in concert with taxonomic labelling can reveal paralogy.
Our conflict analysis tool informs the users of whether their tree contains major conflicts with established taxonomy.
This conflict tool also returns information on if these taxonomic conflicts existed in the original input tree.
These conflicts may be a sign that taxonomy needs to be updated, or may be a sign that non-homologous sequences have been included in the analysis.
These taxonomic and phylogenetic conflicts flag regions for the tree for the researcher to more closely examine and assess homology.

@@

\section*{Conclusions}
Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible Phylogenetics that has the
potential to consistently provide context for these ideas, highlighting the
importance of data sharing and open science in the field, biology and science.
